---
title: "CCBER VMMV"
author: "Joanna Tang"
date: "10/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include = FALSE}

# Load packages, read in data

library(tidyverse)
library(janitor)
library(plotrix)
library(knitr)
library(kableExtra)
library(lubridate)
library(RColorBrewer)



vmmv_monitoring <- read_csv("vmmv_monitoring.csv", col_types = cols(avg_height = col_number(), inflorescences = col_number(), plant_id = col_character()))  %>% mutate(date = as.Date(date, format = "%m/%d/%y"))

vmmv_recruits <- read_csv("vmmv_recruits.csv", col_types = cols(basal_stems = col_number(), date = col_character(), flowering = col_character(), seedling_id = col_number(), series = col_integer(), vigor = col_number())) %>% mutate(date = as.Date(date, format = "%m/%d/%y"))


```

### H1: Zone affects vigor over time.
#### Result: Zone has a significant effect on vigor.  H1 supported.

```{r message = FALSE, warning = FALSE, echo = FALSE}

# H1: Zone affects vigor over time.

## Exploratory graphs
vigor_hist <- ggplot(vmmv_monitoring, aes(x = vigor)) +
  geom_histogram()
#not normal -- skewed right
vigor_qq <- ggplot(vmmv_monitoring, aes(sample = vigor)) +
  geom_qq()
#not normal

# Dataframe calculating median vigor per zone
summary_vigor <- vmmv_monitoring %>% 
  filter(location == "Plover Area") %>% 
  group_by(date, zone) %>% 
  summarize(median_vigor = median(vigor)) %>% 
  drop_na()

# Scatterplot of vigor over time, by zone
vigor_scatter <- vmmv_monitoring %>% 
  filter(location == "Plover Area") %>% 
  filter(zone != "NA") %>% 
  filter(vigor<5) %>% 
  ggplot(aes(x = date, y = as.factor(-vigor), color = zone)) +
  geom_jitter(width = .1) +
  geom_smooth(method = lm, aes(x = date, group = zone, color = zone), se = TRUE, size = .5) +
  labs(title = "Plant vigor over time", x = "Date", y = "Vigor Index", caption = "Graph shows general linear model of plant vigor index (ranging from 1 = healthy to 4 = dead) over time \n Swale has poorest vigor") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_y_discrete(labels = c("4", "3", "2", "1")) +
  scale_x_date(date_labels = "%B 1", breaks = "1 month")
vigor_scatter



# Dataframe calculating % at each vigor
vigor_proportions <- vmmv_monitoring %>% 
  count(date, vigor, zone) %>% 
  group_by(date, vigor, zone) %>% 
  summarize(proportion = sum(n)/30) %>% 
  drop_na() %>% 
  filter(vigor <5)

# Stacked column graph of vigor proportion over time
vigor_col <- vigor_proportions %>% 
  ggplot(aes(x = date, y = proportion, fill = as.factor(vigor))) +
  geom_col(position = "fill") +
  facet_wrap(~zone) +
  scale_fill_manual(values = c("chartreuse4", "yellowgreen", "yellow2", "coral", "brown3", "tan3", "tan4"), name = "Vigor Index") +
  labs(title = "Vigor proportions over time", x = "Date", y = "Proportion", caption = "Graph shows proportion of plant vigor index (ranging from 1 = healthy to 4 = dead) \n Zone has a significant effect on vigor (chi-squared = 59.22, p << 0.001)") +
  theme_classic() +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_date(date_labels = "%B", breaks = "1 month")
vigor_col


## Chi-squared test of vigor proportions at 8/27/20, between zones
vigor_chi_squared_matrix <- vmmv_monitoring %>% 
  filter(date == "2020-08-27") %>% 
  filter(vigor <5) %>% 
  filter(zone != "NA") %>% 
  count(zone, vigor) %>% 
  spread(vigor, n) %>% # tell the fxn which column to spread by and what to fill the obs with
  select(-zone)
vigor_chi_squared_matrix[is.na(vigor_chi_squared_matrix)] <- 0
rownames(vigor_chi_squared_matrix) <- c("Ridge", "Slope", "Swale")
vigor_chi_squared <- chisq.test(vigor_chi_squared_matrix)
#vigor_chi_squared
#p << .001, chi-squared = 59.22

# Table of percentages at each vigor at 8/27/20, between zones
august_vigor_percentages <- kable(prop.table(as.matrix(vigor_chi_squared_matrix), 1) * 100, digits = 2) %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
august_vigor_percentages


```
<i> Table shows percentage of plants at each vigor (ranging from 1 = healthy to 4 = dead) in each zone during August 2020. </i>


### H2: Zone affects fitness over time.
#### Result: Swale plant fitness (average height and average number of inflorescneces) is significantly lower than that of plants in ridge or slope.  H2 supported.

```{r message = FALSE, warning = FALSE, echo = FALSE}

# H2: Zone affects fitness over time.

# Filter out only August monitoring data
august_monitoring <- vmmv_monitoring %>% 
  filter(date == "2020-08-27") %>% 
  mutate(avg_height = as.numeric(avg_height)) %>% 
  mutate(inflorescences = as.numeric(inflorescences))

## Exploratory graphs
august_height_hist <- august_monitoring %>% 
  ggplot() +
  geom_histogram(aes(x = avg_height))
#pretty normal
august_height_qq <- august_monitoring %>% 
  ggplot() +
  geom_qq(aes(sample = avg_height))
#pretty normal
august_infl_hist <- august_monitoring %>% 
  ggplot() +
  geom_histogram(aes(x = inflorescences))
#fairly normal...
august_infl_qq <- august_monitoring %>% 
  ggplot() +
  geom_qq(aes(sample = inflorescences))
#fairly normal...

# Average size and reproduction of plants by zone in August
august_summary <- august_monitoring %>% 
  filter(date == "2020-08-27") %>% 
  filter(avg_height != "NA") %>% 
  filter(inflorescences != "NA") %>% 
  filter(zone != "NA") %>% 
  group_by(zone) %>%
  summarize(mean_height = mean(avg_height),
            se_height = std.error(avg_height),
            mean_infl = mean(inflorescences),
            se_infl = std.error(inflorescences)
            )

# Column graph of August height by zone
august_height_col <- august_summary %>% 
  ggplot(aes(x = zone, y = mean_height)) +
  geom_col() +
  geom_errorbar(aes(x = zone, ymin = mean_height-se_height, ymax = mean_height+se_height, width = .5)) +
  labs(title = "Size of VMMV", x = "Zone", y = "Average height (cm)", caption = "Graph of average height of plants in August 2020, with standard error represented as error bars. \n Average plant height in swale is significantly lower \n than that in ridge and slope (ANOVA F = 12.28, p << 0.01)") +
  theme_classic() +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic()
august_height_col

# August average height ANOVA
august_height_anova <- aov(avg_height ~ zone, data = august_monitoring)
#summary(august_height_anova)
#p << 0.01, F-value = 12.28
#Tukey's post-hoc
august_height_tukey <- TukeyHSD(august_height_anova)
#slope-ridge p = .13, swale-ridge p = .0035, swale-slope p = .000072

# Column graph of August inflorescences by zone
august_infl_col <- august_summary %>% 
  ggplot(aes(x = zone, y = mean_infl)) +
  geom_col() +
  geom_errorbar(aes(x = zone, ymin = mean_infl-se_infl, ymax = mean_infl+se_infl, width = .5)) +
  labs(title = "Reproduction of VMMV", x = "Zone", y = "Average # inflorescences per plant", caption = "Graph of average number of inflorescences per plant in August 2020, with standard error represented as error bars. \n Average number of inflorescneces per plant in swale is significantly lower \n than that in ridge and slope (ANOVA F = 18.33, p << 0.01)") +
  theme_classic() +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic()
august_infl_col

# August average inflorescences ANOVA
august_infl_anova <- aov(inflorescences ~ zone, data = august_monitoring)
#summary(august_infl_anova)
#p << 0.01, F-value = 18.33
#Tukey's post-hoc
august_infl_tukey <- TukeyHSD(august_infl_anova)
#slope-ridge p = .21, swale-ridge p = .0000336, swale-slope p << .001

```

### H3: Zone affects seedling recruitment.
#### Result: Zone has no significant effect on proportion of surviving seedlings in each zone.  H3 rejected.

```{r message = FALSE, warning = FALSE, echo = FALSE}

# H1: Zone affects seedling recruitment.

# Table of # recruits per zone in September 2020
sept_recruits <- vmmv_recruits %>% 
  filter(date == "2020-09-21") %>% 
  filter(zone != "NA") %>% 
  group_by(zone) %>% 
  summarize(recruits = length(zone))

# Dataframe total # recruits per month, per zone
summary_recruits <- vmmv_recruits %>% 
  group_by(date, zone) %>% 
  summarize(sum_seedlings = sum(total_seedlings))

# Line graph of total # recruits per zone over time
recruits_line <- summary_recruits %>% 
  ggplot(aes(x = date, y = sum_seedlings, color = zone)) +
  geom_point() +
  geom_line() +
  labs(title = "Recruitment over time", x = "Date", y = "# seedlings", caption = "Graph shows total number of surviving seedlings over time") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_date(date_labels = "%B 1", breaks = "1 month")
recruits_line

# Column graph of # recruits per zone in September 2020
recruits_col <- sept_recruits %>% 
  ggplot(aes(x = zone, y = recruits)) +
  geom_col() +
  labs(title = "Recruitment", x = "Zone", y = "# seedlings", caption = "Graph shows total number of seedlings in each zone in September 2020 \n No significant differences (chi-squared = 3.82, p = 0.15)") +
  theme_classic() +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic()
recruits_col

# Chi-squared test for September
sept_seedling_chi_squared_matrix <- sept_recruits %>% 
  select(-zone)
sept_seedling_chi_squared_matrix[is.na(sept_seedling_chi_squared_matrix)] <- 0
rownames(sept_seedling_chi_squared_matrix) <- c("Ridge", "Slope", "Swale")
sept_seedling_chi_squared <- chisq.test(sept_seedling_chi_squared_matrix)
#sept_seedling_chi_squared
#p = .1484, chi-squared = 3.854


# Dataframe of # seedlings per plant, per zone
summary_recruits_per_plant <- vmmv_recruits %>% 
  group_by(parent_plant_id, date, zone) %>% 
  summarize(sum_seedlings = sum(total_seedlings)) %>% 
  group_by(date, zone) %>% 
  summarize(mean_seedlings_per_plant = mean(sum_seedlings),
            se = std.error(sum_seedlings))
  

# Line graph of # seedlings per plant, per zone over time
recruits_per_plant_line <- summary_recruits_per_plant %>% 
  ggplot(aes(x = date, y = mean_seedlings_per_plant, color = zone)) +
  geom_point() +
  geom_line() +
  labs(title = "Recruitment per plant over time", x = "Date", y = "# seedlings per plant", caption = "Graph shows average number of seedlings per plant over time") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_date(date_labels = "%B 1", breaks = "1 month")
recruits_per_plant_line

# Column graph of # recruits per plant, per zone in September 2020
recruits_per_plant_col <- summary_recruits_per_plant %>% 
  filter(date == "2020-09-21") %>%
  ggplot(aes(x = zone, y = mean_seedlings_per_plant)) +
  geom_col() +
  geom_errorbar(aes(x = zone, ymin = mean_seedlings_per_plant-se, ymax = mean_seedlings_per_plant+se, width = .5)) +
  labs(title = "Recruitment per plant", x = "Zone", y = "# seedlings per plant", caption = "Graph shows average number of seedlings per plant in each zone in September 2020, \n with error bars representing standard error \n No significant differences (Kruskal-Wallis chi-squared = 0.14, p = 0.93)") +
  theme_classic() +
  scale_y_continuous(expand = c(0,0)) +
  scale_fill_grey(name = "Lineage") +
  theme_classic()
recruits_per_plant_col


# Exploratory graphs
sept_avg_recruits_hist <- vmmv_recruits %>% 
  group_by(parent_plant_id, date, maternal_accession) %>% 
  filter(maternal_accession != "NA") %>% 
  filter(date == "2020-09-21") %>% 
  summarize(sum_seedlings = sum(total_seedlings)) %>% 
  ggplot(aes(x = sum_seedlings)) +
  geom_histogram()
#skewed right
sept_avg_recruits_qq <- vmmv_recruits %>% 
  group_by(parent_plant_id, date, maternal_accession) %>% 
  filter(maternal_accession != "NA") %>% 
  filter(date == "2020-09-21") %>% 
  summarize(sum_seedlings = sum(total_seedlings)) %>% 
  ggplot(aes(sample = sum_seedlings)) +
  geom_qq()
#skewed right

# Kruskal-Wallis for September # seedlings per plant, by zone
sept_seedling_per_plant_zone_df <- vmmv_recruits %>% 
  group_by(parent_plant_id, date, zone) %>% 
  summarize(sum_seedlings = sum(total_seedlings)) %>% 
  filter(date == "2020-09-21")
sept_seedling_per_plant_zone_kw <- kruskal.test(as.factor(sum_seedlings) ~ as.factor(zone), data = sept_seedling_per_plant_zone_df)
#p = .9327, chi-squared = .13924, df = 2




```

### H4: Maternal accession has no effect on seedling recruitment.
#### Result: Maternal accession has no significant effect on proportion of surviving seedlings per plant from each accession.  H4 supported.

```{r message = FALSE, warning = FALSE, echo = FALSE}

# H4: Maternal accession has no effect on seedling recruitment.

# Dataframe total # recruits per month, per accession & lineage
accessions_summary_recruits <- vmmv_recruits %>% 
  group_by(date, maternal_accession) %>% 
  filter(lineage != "NA") %>% 
  summarize(sum_seedlings = sum(total_seedlings))

# Line graph of total # recruits per zone over time
accession_recruits_line <- accessions_summary_recruits %>% 
  ggplot(aes(x = date, y = sum_seedlings, color = maternal_accession)) +
  geom_point() +
  geom_line() +
  labs(title = "Recruitment over time", x = "Date", y = "# seedlings", caption = "Graph shows total number of surviving seedlings over time") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_color_brewer(palette = "Paired", name = "Maternal Accession") +
  scale_x_date(date_labels = "%B 1", breaks = "1 month")
accession_recruits_line

# Column graph of # recruits per maternal accession in September 2020
accession_recruits_col <- accessions_summary_recruits %>% 
  filter(date == "2020-09-21") %>%
  ggplot(aes(x = maternal_accession, y = sum_seedlings)) +
  geom_col() +
  labs(title = "Recruitment", x = "Maternal Accession", y = "# seedlings", caption = "Graph shows total number of seedlings from each accession in September 2020 \n Maternal accession has significant effect on total number of seedlings (chi-squared = 18.13, p = 0.020)") +
  theme_classic() +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic()
accession_recruits_col

# Chi-squared test for September
accession_sept_seedling_chi_squared_matrix <- accessions_summary_recruits %>% 
  filter(date == "2020-09-21") %>% 
  group_by(maternal_accession) %>% 
  summarize(sum = sum(sum_seedlings)) %>% 
  select(-maternal_accession)
accession_sept_seedling_chi_squared_matrix[is.na(accession_sept_seedling_chi_squared_matrix)] <- 0
rownames(accession_sept_seedling_chi_squared_matrix) <- c("A2051", "A3002", "A3003", "A3015", "A3041", "A3051", "B227", "B29R", "B4001")
accession_sept_seedling_chi_squared <- chisq.test(accession_sept_seedling_chi_squared_matrix)
#accession_sept_seedling_chi_squared
#p = .02028, chi-squared = 18.129


# Dataframe of # seedlings per plant, per maternal accession
accessions_recruits_per_plant <- vmmv_recruits %>% 
  group_by(parent_plant_id, date, maternal_accession) %>% 
  filter(maternal_accession != "NA") %>% 
  summarize(sum_seedlings = sum(total_seedlings)) %>% 
  group_by(date, maternal_accession) %>% 
  summarize(mean_seedlings_per_plant = mean(sum_seedlings))


# Line graph of # seedlings per plant, per zone over time
accessions_recruits_per_plant_line <- accessions_recruits_per_plant %>% 
  ggplot(aes(x = date, y = mean_seedlings_per_plant, color = maternal_accession)) +
  geom_point() +
  geom_line() +
  labs(title = "Recruitment per plant over time", x = "Date", y = "# seedlings per plant", caption = "Graph shows average number of seedlings per plant over time") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_color_brewer(palette = "Paired", name = "Maternal Accession") +
  scale_x_date(date_labels = "%B 1", breaks = "1 month")
accessions_recruits_per_plant_line

# Column graph of # recruits per plant, per maternal accession in September 2020
accession_recruits_per_plant_col <- vmmv_recruits %>% 
  filter(date == "2020-09-21") %>%
  group_by(parent_plant_id, date, maternal_accession, lineage) %>% 
  filter(maternal_accession != "NA") %>% 
  summarize(sum_seedlings = sum(total_seedlings)) %>% 
  group_by(date, maternal_accession) %>% 
  summarize(mean_seedlings_per_plant = mean(sum_seedlings),
            se = std.error(sum_seedlings)) %>% 
  ggplot(aes(x = maternal_accession, y = mean_seedlings_per_plant)) +
  geom_col() +
  geom_errorbar(aes(x = maternal_accession, ymin = mean_seedlings_per_plant-se, ymax = mean_seedlings_per_plant+se, width = .5)) + 
  labs(title = "Recruitment per plant", x = "Maternal Accession", y = "# seedlings per plant", caption = "Graph shows average number of seedlings per plant from each lineage in September 2020 \n No significant differences (Kruskal-Wallis chi-squared = 7.57, p = 0.48)") +
  theme_classic() +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic()
accession_recruits_per_plant_col


# Kruskall-Wallis for September # seedlings per plant, by accession
sept_seedling_per_plant_accession_df <- vmmv_recruits %>% 
  group_by(parent_plant_id, date, maternal_accession) %>% 
  summarize(sum_seedlings = sum(total_seedlings)) %>% 
  filter(date == "2020-09-21")
sept_seedling_per_plant_accession_kw <- kruskal.test(as.factor(sum_seedlings) ~ as.factor(maternal_accession), data = sept_seedling_per_plant_accession_df)
#p = .4761, chi-squared = 7.5741, df = 8





```


### H5: Maternal accession has no effect on vigor.
#### Result: Maternal accession has no significant effect on plant vigor.  H5 supported.

```{r message = FALSE, warning = FALSE, echo = FALSE}

# H5: Maternal accession has no effect on vigor.

# Dataframe of vigor over time, by maternal accession
accession_summary_vigor <- vmmv_monitoring %>% 
  filter(location == "Plover Area") %>% 
  group_by(date, maternal_accession) %>% 
  summarize(median_vigor = median(vigor)) %>% 
  drop_na()

# Scatterplot of vigor over time, by zone
accession_vigor_scatter <- vmmv_monitoring %>% 
  filter(location == "Plover Area") %>% 
  filter(maternal_accession != "NA") %>% 
  filter(vigor<5) %>% 
  ggplot(aes(x = date, y = as.factor(-vigor), color = maternal_accession)) +
  geom_jitter(width = .1) +
  geom_smooth(method = lm, aes(x = date, group = maternal_accession, color = maternal_accession), se = TRUE, size = .5) +
  labs(title = "Plant vigor over time", x = "Date", y = "Vigor Index", caption = "Graph shows general linear model of plant vigor index (ranging from 1 = healthy to 4 = dead) for each accession") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_y_discrete(labels = c("4", "3", "2", "1")) +
  scale_x_date(date_labels = "%B 1", breaks = "1 month") +
  scale_color_brewer(palette = "Paired", name = "Maternal Acceession")
#accession_vigor_scatter



# Dataframe calculating total # plants from each accession at each monitoring date
accession_totals <- vmmv_monitoring %>% 
  filter(maternal_accession != "NA") %>% 
  count(date, vigor, maternal_accession) %>% 
  group_by(date, maternal_accession) %>% 
  summarize(total_per_accession = sum(n))

# Dataframe calculating % at each vigor, per accession
accession_vigor_proportions <- vmmv_monitoring %>% 
  full_join(accession_totals) %>% 
  filter(maternal_accession != "NA") %>% 
  count(date, vigor, maternal_accession, total_per_accession) %>% 
  mutate(proportion = n/total_per_accession) %>% 
  drop_na() %>% 
  filter(vigor <5)

# Stacked column graph of vigor proportion over time, per accession
accession_vigor_col <- accession_vigor_proportions %>% 
  ggplot(aes(x = date, y = proportion, fill = as.factor(vigor))) +
  geom_col(position = "fill") +
  facet_wrap(~maternal_accession) +
  scale_fill_manual(values = c("chartreuse4", "yellowgreen", "yellow2", "coral", "brown3", "tan3", "tan4"), name = "Vigor Index") +
  labs(title = "Vigor proportions over time", x = "Date", y = "Proportion", caption = "Graph shows proportion of plant vigor index (ranging from 1 = healthy to 4 = dead) \n Maternal acceession has no effect on vigor (chi-squared = 30.20, p = 0.18)") +
  theme_classic() +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_date(date_labels = "%B", breaks = "1 month")
accession_vigor_col


## Chi-squared test of vigor proportions at 8/27/20, between accessions
accession_vigor_chi_squared_matrix <- vmmv_monitoring %>% 
  filter(date == "2020-08-27") %>% 
  filter(vigor <5) %>% 
  filter(maternal_accession != "NA") %>% 
  count(maternal_accession, vigor) %>% 
  spread(vigor, n) %>% # tell the fxn which column to spread by and what to fill the obs with
  select(-maternal_accession)
accession_vigor_chi_squared_matrix[is.na(accession_vigor_chi_squared_matrix)] <- 0
rownames(accession_vigor_chi_squared_matrix) <- c("A2051", "A3002", "A3003", "A3015", "A3041", "A3051", "B227", "B29R", "B4001")
accession_vigor_chi_squared <- chisq.test(accession_vigor_chi_squared_matrix)
#accession_vigor_chi_squared
#p = .1783, chi-squared = 30.199

# Table of percentages at each vigor at 8/27/20, between accessions
august_accession_vigor_percentages <- kable(prop.table(as.matrix(accession_vigor_chi_squared_matrix), 1) * 100, digits = 2) %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
august_accession_vigor_percentages




```
<i> Table shows percentage of plants at each vigor (ranging from 1 = healthy to 4 = dead) from each accession during August 2020. </i>

